---
title: "Estimating TRAs from GTex"
author: "Hannah Meyer"
date: "`r Sys.Date()`"
output:
    pdf_document:
        fig_caption: yes
        toc: true
        toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=FALSE, message=FALSE)
```

## Tissue-restricted antigens
During T cell development in the thymus, thymocytes encounter more then 85% of
all protein, including proteins that are usually expressed in a tissue specific
manner. In the following, we use a gene expression data set of about 50k
transcripts across 30 tissues (GTEx v8) to estimate tissue specific genes.
We use the measure $\tau$ as introduced in
[Yanai 2005](https://academic.oup.com/bioinformatics/article/21/5/650/220059)
and benchmarked in
[Kryuchkova-Mostacci 2017](https://academic.oup.com/bib/article/18/2/205/2562739#119555168)
to determine a list of human tissue-restricted antigens:

$$\tau=\frac{\sum_{i=1}^{n}(1-\hat{x}_i)}{n-1}$$
with
$$\hat{x_i} = \frac{x_i}{\max\limits_{1\leq i\leq n}{x_i}}$$
where $x_i$ is the expression of the gene in tissue $i$ and $n$ is the number of
tissues.

## Setup
```{r setup variables and R}
library(tidyverse)
box::use(tra=./tra)
datadir <- "~/data/tra/gtex"
text_size <- 7
title_size <- 9
```

## Download files from GTex: all genes quantified

```{bash download files, eval=FALSE}
gtex=https://storage.googleapis.com/gtex_analysis_v8/annotations
datadir=~/data/tra/gtex

cd $datadir
wget $gtex/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt
wget $gtex/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz
gunzip $gtex/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz
```

The following description of the  Quality Control and gene expression analysis
is taken 1:1 from the [GTEx website](https://gtexportal.org/home/documentationPage#staticTextAnalysisMethods)

### RNA-seq Alignment
Alignment to the human reference genome hg19/GRCh37 was performed using STAR
v2.4.2a, based on the GENCODE v19 annotation. Unaligned reads were kept in the
final BAM file. Among multi-mapping reads, one read is flagged as the primary
alignment by STAR.

### Quantification

Gene-level quantifications: read counts and TPM values were produced with
RNA-SeQC v1.1.8 (DeLuca et al., Bioinformatics, 2012), using the following
read-level filters:

1. Reads were uniquely mapped (corresponding to a mapping quality of 255 for
   START BAMs).
1. Reads were aligned in proper pairs.
1. The read alignment distance was <= 6 (i.e., alignments must not contain more
   than six non-reference bases).
1. Reads were fully contained within exon boundaries. Reads overlapping introns
   were not counted.
These filters were applied using the "-strictMode" flag in RNA-SeQC.

### QC and Sample Exclusion Process

1. RNA-seq expression outliers were identified and excluded using a
   multidimensional extension of the statistic described in (Wright et al.,
   Nat. Genet. 2014 ). Briefly, for each tissue, read counts from each sample
   were normalized using size factors calculated with DESeq2 and log-transformed
   with an offset of 1; genes with a log-transformed value >1 in >10% of samples
   were selected, and the resulting read counts were centered and unit-
   normalized. The resulting matrix was then hierarchically clustered (based on
   average and cosine distance), and a chi2 p-value was calculated based on
   Mahalanobis distance. Clusters with >=60% samples with Bonferroni-corrected
   p-values <0.05 were marked as outliers, and their samples were excluded.
1. Samples with <10 million mapped reads were removed.
1. For samples with replicates, the replicate with the greatest number of reads
   was selected.
   
### Expression analysis

Gene expression values for all samples from a given tissue were normalized using
the following procedure:

1. Genes were selected based on expression thresholds of > 0.1 TPM in at least
   20% of samples and >=6 reads in at least 20% of samples.
1. Expression values were normalized between samples using TMM as implemented in
   edgeR (Robinson & Oshlack, Genome Biology, 2010).
1. For each gene, expression values were normalized across samples using an
   inverse normal transform.

## Load gene expression data sets
```{r load data, eval=FALSE}
description_attr <- data.table::fread(file.path(datadir,
                                "GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt"),
                                data.table=FALSE) %>%
                    as_tibble

samples_tpm <- data.table::fread(file.path(datadir,
                                "GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct"),
                                data.table=FALSE) %>%
                  as_tibble

```
## Format gene expression data sets
Join gene ids, tissue and sample IDs with expression values.
```{r annotate tpm, eval=FALSE}
tpm_attr <- samples_tpm %>%
    select(Name, Description) %>%
    mutate(ID=gsub("\\..*", "", Name))

write_csv(tpm_attr,
          file.path(datadir,
                    "GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm_attr.csv")
)

tpm_annotated <- samples_tpm %>%
    select(-Description, -Name) %>%
    t %>%
    magrittr::set_colnames(tpm_attr$Name) %>%
    as_tibble %>% 
    mutate(SAMPID=colnames(samples_tpm)[-c(1:2)]) %>% 
    inner_join(select(description_attr, SAMPID, SMTS), by = "SAMPID")

write_csv(tpm_annotated,
          file.path(datadir,
                    "GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm_gct_annotated.csv")
)
```

Compute the mean expression per gene and per tissue across biological replicates
```{r mean expression, eval=FALSE}
mean_expression <- tpm_annotated %>%
    select(-SAMPID) %>%
    group_by(SMTS) %>%
    summarise_all(mean, na.rm=TRUE)

tissues_cols <- mean_expression$SMTS
mean_expression <- t(mean_expression[,-1])
colnames(mean_expression) <- tissues_cols
saveRDS(mean_expression, "GTEx_Analysis_v8_gene_mean_expression.Rdata")
```

##  GTex data: only refseq genes expressed in xx tissues

```{r eval=FALSE}
gtex_genes <- read_csv(file.path(datadir,"GTEx_Genes.csv"))
colnames(gtex_genes)[1] <- "ID"
gtex_genes <- gtex_genes %>%
    drop_na(ID) %>%
    column_to_rownames("ID") %>%
    t %>%
    as.data.frame %>%
    rownames_to_column("ID") %>%
    mutate(tissue = gsub("(.*)_GTEX.*", "\\1", ID))

gtex_genes_mean <- gtex_genes %>%
    select(-ID) %>%
    group_by(tissue) %>%
    summarize(across(.fns=mean)) %>%
    ungroup %>%
    column_to_rownames("tissue") %>%
    t

gtex_transcripts <- read_csv(file.path(datadir,"GTEx_Transcripts.csv"))
colnames(gtex_transcripts)[1] <- "ID"
gtex_transcripts <- gtex_transcripts  %>%
    drop_na(ID) %>%
    column_to_rownames("ID") %>%
    t %>%
    as.data.frame %>%
    rownames_to_column("ID") %>%
    mutate(tissue = gsub("(.*)_GTEX.*", "\\1", ID))
gtex_transcripts_mean <- gtex_transcripts %>%
    select(-ID) %>%
    group_by(tissue) %>%
    summarize(across(.fns=mean)) %>%
    ungroup %>%
    column_to_rownames("tissue") %>%
    t
```

```{r, echo=FALSE, eval=FALSE}
saveRDS(gtex_transcripts_mean, file.path(datadir, "gtex_transcripts_mean.Rdata"))
saveRDS(gtex_genes_mean, file.path(datadir, "gtex_genes_mean.Rdata"))
```



```{r, echo=FALSE}
tpm_attr <- read_csv(file.path(datadir,
                    "GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm_attr.csv"))
gtex_all_mean <- readRDS("GTEx_Analysis_v8_gene_mean_expression.Rdata")
gtex_transcripts_mean <- readRDS(file.path(datadir, "gtex_transcripts_mean.Rdata"))
gtex_genes_mean <- readRDS(file.path(datadir, "gtex_genes_mean.Rdata"))
```

## Compute list of tissue-specific genes
We follow the processing processing procedure described in the
[Kryuchkova-Mostacci 2017](https://academic.oup.com/bib/article/18/2/205/2562739#119555168)
benchmarking study:

* All genes where the highest gene expression < 1 TPM are set as not expressed;
* count data are log-transformed after addition of 1 to avoid zero counts.

```{r tau gtex}
filter_expression <-  function(mean_expression,  tpm=1) {
    
    gtex_max <- apply(mean_expression, 1, tra$fmax)
    mean_expression_filter <- mean_expression[gtex_max > tpm,]
    mean_expression_filter <- mean_expression_filter + 1
    
    return(mean_expression_filter)
}

get_tra <- function(mean_expression_filter, study, tpm=1) {
    
    gtex <- data.frame(id=gsub("\\..*", "", rownames(mean_expression_filter)),
                       full=rownames(mean_expression_filter),
                       tau=apply(log2(mean_expression_filter), 1, tra$tau),
                       study=study,
                       stringsAsFactors = FALSE)
    
    write.table(select(gtex, -study),
            file=file.path(datadir, paste0(study, "_tau.csv")),
            sep=",", col.names=TRUE, row.names=FALSE, quote=FALSE)
    
    tra <- gtex %>% 
        select(-study) %>%
        filter(tau >= 0.8)
    
    write.table(tra,
            file=file.path(datadir, paste0(study, "_tra_tau_ge_0.8.csv")),
            sep=",", col.names=TRUE, row.names=FALSE, quote=FALSE)
    
    return(gtex)
}

gtex_genes_filter <- filter_expression(gtex_genes_mean)
gtex_transcripts_filter <- filter_expression(gtex_transcripts_mean)
gtex_all_filter <- filter_expression(gtex_all_mean)

gtex_genes_tau <- get_tra(gtex_genes_filter, "gtex_genes")
gtex_transcripts_tau <- get_tra(gtex_transcripts_filter , "gtex_transcripts")
gtex_all_tau <- get_tra(gtex_all_filter, "gtex_genes_all")

gtex_combine <- rbind(gtex_genes_tau,gtex_transcripts_tau, gtex_all_tau)
p <- ggplot(gtex_combine, aes(x=tau))
p + geom_density(aes(color=study)) +
    scale_color_brewer(type='qual', palette = "Set1") +
    cowplot::theme_cowplot() +
    theme(axis.text = element_text(size=text_size),
          axis.title = element_text(size=title_size))

```


```{r}
tt <- tibble(GTEx = c("All GTEx Genes", "RefSeq Genes", "RefSeq Transcripts"),
       total= c(nrow(gtex_all_mean), nrow(gtex_genes_mean), nrow(gtex_transcripts_mean)),
       expressed = c(nrow(gtex_all_tau),
                                nrow(gtex_genes_tau),
                                nrow(gtex_transcripts_tau)),
       tra_0.7 = c(sum(gtex_all_tau$tau >= 0.7),
                            sum(gtex_genes_tau$tau >= 0.7),
                            sum(gtex_transcripts_tau$tau >= 0.7)),
       tra_0.8 = c(sum(gtex_all_tau$tau >= 0.8),
                            sum(gtex_genes_tau$tau >= 0.8),
                            sum(gtex_transcripts_tau$tau >= 0.8))
)
kableExtra::kable(tt, format="latex",
                  col.names = c("GTEx", "Total", "Expressed: TPM $\\ge$ 1",
                                "TRA: $\\tau \\ge$ 0.7",
                                "TRA: $\\tau \\ge$ 0.8"))
```

## Find tissue-specificity of tissue-specific genes
* use binarizing approach described by 
[Yanai 2005](https://academic.oup.com/bioinformatics/article/21/5/650/220059)

```{r}
find_tissue_spec <- function(mean_expression_filter, feature_tau, study,
                             tauthr=0.8) {
    feature_tau <- feature_tau[feature_tau$tau > tauthr,]
    mean_expression_tra <- mean_expression_filter %>%
        as.data.frame %>%
        rownames_to_column("full") %>%
        filter(full %in% feature_tau$full) %>%
        column_to_rownames("full") %>%
        as.matrix

    expression_tra_tissues <- apply(mean_expression_tra, 1,
                                    tra$find_tra_tissues) %>%
        t %>%
        as.data.frame %>%
        rownames_to_column("full") %>%
        left_join(feature_tau) %>%
        select(-full, -study) %>%
        select(id, tau, everything()) %>%
        as_tibble()

    write.table(expression_tra_tissues,
            file=file.path(datadir, paste0(study, "_binarized_expression.csv")),
            sep=",", col.names=TRUE, row.names=FALSE, quote=FALSE)
    
    return(expression_tra_tissues)
}

gtex_genes_tissues <- find_tissue_spec(gtex_genes_filter, gtex_genes_tau,
                                       "gtex_genes")
gtex_transcripts_tissues <- find_tissue_spec(gtex_transcripts_filter, gtex_transcripts_tau,
                                       "gtex_transcripts")
gtex_genes_all_tissues <- find_tissue_spec(gtex_all_filter, gtex_all_tau,
                                       "gtex_genes_all")

```

* visualise distribution of TRAs per tissue

```{r fig.keep="hold"}
visualise_tra_per_tissue <- function(expression_tra_tissues) {
    tras_per_tissue <- expression_tra_tissues %>%
        mutate(thr=case_when(tau < 0.8 ~ "> 0.7", 
                             tau >= 0.8 & tau < 0.9 ~ "> 0.8",
                             tau >= 0.9 ~ "> 0.9")) %>%
        pivot_longer(-c(id, tau, thr),
                     names_to="tissue", values_to="status") %>%
        filter(status != 0) %>%
        group_by(tissue, thr) %>%
        summarise(tras = n(), .groups='drop') %>%
        arrange(tras) %>%
        mutate(tissue = fct_inorder(tissue))
    
    p_all_tissues <- ggplot(tras_per_tissue) +
        geom_bar(aes(x=tissue, y=tras, fill=thr), stat='identity') +
        scale_fill_manual(values=c('#66c2a5','#fc8d62','#8da0cb'), guide=FALSE) +
        labs(x="GTEx tissues",
             y="Number of TRAs",
             fill="Tau threshold") +
        cowplot::theme_cowplot() +
        theme(axis.text.x = element_blank(),
              axis.title.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.text.y = element_text(size=text_size),
              axis.title.y = element_text(size=title_size))
    
    p_tissues_no_testes <- ggplot(filter(tras_per_tissue, tissue != "Testis")) +
        geom_bar(aes(x=tissue, y=tras, fill=thr), stat='identity') +
        scale_fill_manual(values=c('#66c2a5','#fc8d62','#8da0cb')) +
        labs(x="GTEx tissues",
             y="Number of TRAs",
             fill="Tau threshold") +
        cowplot::theme_cowplot() +
        theme(axis.text.x = element_text(angle=45, hjust = 1, vjust = 1),
              axis.text = element_text(size=text_size),
              axis.title = element_text(size=title_size),
              legend.position = "bottom")
    cowplot::plot_grid(p_all_tissues, p_tissues_no_testes,
                       nrow=2,
                       align="v",
                       rel_heights = c(1, 2.5),
                       axis="lr"
                       )
}
gtex_genes_tra_per_tissue <- visualise_tra_per_tissue(gtex_genes_tissues)
gtex_transcripts_tra_per_tissue <- visualise_tra_per_tissue(gtex_transcripts_tissues)
gtex_all_tra_per_tissue <- visualise_tra_per_tissue(gtex_genes_all_tissues)

gtex_genes_tra_per_tissue
gtex_transcripts_tra_per_tissue
gtex_all_tra_per_tissue
```

* visualize how many tissues a given TRA is assigned
* visualize relationship between tau and number of tissues expressing TRA

```{r fig.keep="hold"}
visualise_tissue_per_tra <- function(expression_tra_tissues) {
    tissue_per_tra <- expression_tra_tissues %>%
        mutate(thr=case_when(tau < 0.8 ~ "> 0.7", 
                             tau >= 0.8 & tau < 0.9 ~ "> 0.8",
                             tau >= 0.9 ~ "> 0.9")) %>%
        pivot_longer(-c(id, tau, thr), names_to="tissue", values_to="status") %>%
        filter(status != 0) %>%
        group_by(id, tau, thr) %>%
        summarise(tissues = n(), .groups='drop') %>%
        arrange(tissues) %>%
        mutate(id = fct_inorder(id))
    
    p_all_tras <-  ggplot(tissue_per_tra) +
        geom_bar(aes(tissues, fill=thr)) +
        scale_fill_manual(values=c('#66c2a5','#fc8d62','#8da0cb')) +
        scale_x_continuous( breaks=1:6)+
        labs(x="Number of GTEx tissues",
             y="Number of TRAs",
             fill="Tau threshold") +
        cowplot::theme_cowplot() +
        theme(axis.text = element_text(size=text_size),
              axis.title = element_text(size=title_size))
    
    p_tau_versus_tissue <-  ggplot(tissue_per_tra) +
        geom_boxplot(aes(x=as.factor(tissues), y=tau, color=thr)) +
        scale_color_manual(values=c('#66c2a5','#fc8d62','#8da0cb')) +
        labs(x="Number of GTEx tissues",
             y="Tau",
             color="Tau threshold") +
        cowplot::theme_cowplot() +
        theme(axis.text = element_text(size=text_size),
              axis.title = element_text(size=title_size)) 
    
    cowplot::plot_grid(p_all_tras,p_tau_versus_tissue,
                       nrow=2,
                       align="v",
                       axis="lr")
}

gtex_genes_tissue_per_tra <- visualise_tissue_per_tra(gtex_genes_tissues)
gtex_transcripts_tissue_per_tra  <- visualise_tissue_per_tra(gtex_transcripts_tissues)
gtex_all_tissue_per_tra  <- visualise_tissue_per_tra(gtex_genes_all_tissues)

gtex_genes_tissue_per_tra
gtex_transcripts_tissue_per_tra
gtex_all_tissue_per_tra
```

## Compare to Fagerberg (2014) data
* Data set derived from Fagerberg (2014) Supplementary data;
* formatting in analogy to formating in Mostacci (2014);
* formatting for zero counts adjusted to adding 1 to each count value instead of
only setting 0 counts to 1.

```{r fagerberg data}
fb_file<- "~/data/public/2016_bioinformatics_mostacci/2014_MCP_Fagerberg.xlsx"
orgExpression <- readxl::read_xlsx(fb_file)
colnames(orgExpression)[1] <- "Ensembl.Gene.ID"

orgExpression <-
    orgExpression[regexpr("ENS", orgExpression$Ensembl.Gene.ID) > 0 |
                      regexpr("FBgn", orgExpression$Ensembl.Gene.ID) > 0 |
                      regexpr("PPAG", orgExpression$Ensembl.Gene.ID) > 0, ]
orgExpression <- na.omit(orgExpression[,-29])
x <- orgExpression[,-1] + 1
orgExpression[, -1] <- log2(x)
fagerberg_max <- apply(orgExpression[,-1], 1, tra$fmax)
orgExpression <- orgExpression[fagerberg_max > log2(1),]
fagerberg <- data.frame(id=orgExpression$Ensembl.Gene.ID,
                        full=orgExpression$Ensembl.Gene.ID,
                        tau=apply(orgExpression[,-1], 1, tra$tau),
                        study='fagerberg',
                        stringsAsFactors = FALSE)

fagerberg <- fagerberg %>% 
    left_join(tpm_attr, by = c("id" = "ID")) %>%
    select(Description, everything()) %>%
    select(-Name) %>%
    rename(Gene=Description)

p <- ggplot(fagerberg, aes(x=tau))
p + geom_density(color="#1b9e77") +
    cowplot::theme_cowplot() +
    theme(axis.text = element_text(size=text_size),
          axis.title = element_text(size=title_size))
```

## Compare the results of GTEx and Fagerberg tissue specificity
* Fagerberg `r nrow(fagerberg)` genes;
* GTEx (all genes: `r nrow(gtex_all_filter)`, refseq only genes 
`r nrow(gtex_genes_filter)`);
* Find common genes and compare distribution of tau;
* check overlap of TRA sets

```{r compare}
all_ids <-  c(fagerberg$id, gtex_all_tau$id, gtex_genes_tau$id)
common <-  table(all_ids)[table(all_ids) == 3]

fagerberg_common <- fagerberg[fagerberg$id %in% names(common),]
gtex_all_genes_common <- gtex_all_tau[gtex_all_tau$id %in% names(common),]
gtex_genes_common <- gtex_genes_tau[gtex_genes_tau$id %in% names(common),]

combined <- rbind(fagerberg_common[, -1], gtex_all_genes_common, gtex_genes_common)

p <- ggplot(combined, aes(x=tau, color=study))
p + geom_density() +
    scale_color_manual(values=c("#1b9e77", "#e41a1c", "#377eb8"))+
    cowplot::theme_cowplot() +
    theme(axis.text = element_text(size=text_size),
          axis.title = element_text(size=title_size),
          legend.position = 'bottom')

gtex_all_genes_tissue_specific <- gtex_all_genes_common[gtex_all_genes_common$tau > 0.8,]
gtex_genes_tissue_specific <- gtex_genes_common[gtex_genes_common$tau > 0.8,]
fagerberg_tissue_specific <- fagerberg_common[fagerberg_common$tau > 0.8,]

gtex_all_genes_pct <- sum(gtex_all_genes_tissue_specific$id %in% fagerberg_tissue_specific$id)/
    nrow(gtex_all_genes_tissue_specific)

gtex_genes_pct <- sum(gtex_genes_tissue_specific$id %in% fagerberg_tissue_specific$id)/
    nrow(gtex_genes_tissue_specific)

fagerberg_all_genes_pct <- sum(gtex_all_genes_tissue_specific$id %in%
                         fagerberg_tissue_specific$id)/
    nrow(fagerberg_tissue_specific)

fagerberg_genes_pct <- sum(gtex_genes_tissue_specific$id %in%
                         fagerberg_tissue_specific$id)/
    nrow(fagerberg_tissue_specific)


```
There are `r length(common)` common genes in the two GTEx (all genes,
refseq genes) and Fagerberg dataset.
Of those, there are `r nrow(gtex_all_genes_tissue_specific)`  and
`r nrow(gtex_genes_tissue_specific)` genes with $\tau > 0.8$ in
the GTEx all genes and refseq genes datasets, respectively, ie genes we
consider as TRA, and
`r nrow(fagerberg_tissue_specific)` TRAs in the Fagerberg dataset. The overlap
between these GTEx and Fagerberg datasets are `r round(gtex_all_genes_pct,2)`
and `r round(fagerberg_all_genes_pct,2)` for all genes in GTEx and
`r round(gtex_genes_pct,2)` and `r round(fagerberg_genes_pct,2)` for
refseq only genes.



