---
title: "TRA_sansom"
author: "Hannah Meyer"
date: "13/06/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Reproduction of dynamic step tissue specificity calls
##Get the data
Download and read in the gcrma normalised GNF3 (GSE10246) expression data, the
GNF3 sample grouping information, annotation (as used for the GR paper) and the
tissue specificity calls as extracted from supplementary table 2

```{r download}
download.file("https://dl.dropboxusercontent.com/s/ro9h4hee2rotzpd/gnf3_gcrma_norm_exprs_matrix.txt.gz?dl=0",
              destfile = "gnf3_gcrma_norm_exprs_matrix.txt.gz")
download.file("https://dl.dropboxusercontent.com/s/dycjzmqjjjq5w3l/gnf3_sample_index.txt.gz?dl=0",
              destfile = "gnf3_sample_index.txt.gz")
download.file("https://dl.dropboxusercontent.com/s/gecb4f31uqt41mi/biomart_ensembl_annotation.txt.gz?dl=0",
              destfile = "biomart_ensembl_annotation.txt.gz")
download.file("https://dl.dropboxusercontent.com/s/1rqhnto22fh6tvy/gr_paper_supp_table2_tse_calls.txt.gz?dl=0",
              destfile = "gr_paper_supp_table2_tse_calls.txt.gz")

gnf3_data <- read.table(gzfile("gnf3_gcrma_norm_exprs_matrix.txt.gz"),
                        sep="\t", header=T, as.is=T)
gnf3_sample_index <- read.table(gzfile("gnf3_sample_index.txt.gz"),
                      sep="\t", header=T, as.is=T)
annotation <- read.table(gzfile("biomart_ensembl_annotation.txt.gz"),
                         header=T, sep="\t", as.is=T)
gr_paper_tse_calls <- read.table(gzfile("gr_paper_supp_table2_tse_calls.txt.gz"),
                                 sep="\t", header=T, as.is=T)
```

## Score the microarray probes for tissue specificity in the tissue groups

```{r}
# Make a dataframe containing the maximum expression values per group
group_expression <- data.frame(row.names=gnf3_data$ID_REF)
for(group in unique(gnf3_sample_index$Tissue.Group[gnf3_sample_index$Include==1])) {
 group_expression[[group]] <- apply(
   gnf3_data[,gnf3_sample_index$Sample_geo_accession[gnf3_sample_index$Tissue.Group == group]],
   1,
   max)
}
# Define a function to call tissue restricted expression based on the
# dynamic step method
dynamic_step <- function(x) {
x <- 2^x # untransform the data
x <- rev(sort(x)) # sort groups from high to low
y <- (x[6]^2 / 50) + x[6] + 50 # compute the per-probe threshold for TRE
if(x[1] > y) {
   return("tissue_restricted")
 } else if (max(x) >= 50) {
    return("not_tissue_restricted")
   } else {
     return("not_detected")
 }
}
# Make the calls
calls <- apply(group_expression,1,dynamic_step)
```

## Compare the results to those published with the GR paper
```{r gr paper}
tre_probes <- names(calls)[calls=="tissue_restricted"]
non_tre_probes <- names(calls)[calls=="not_tissue_restricted"]
non_tre_genes <- unique(annotation$mgi_symbol[annotation$affy_probeset %in% non_tre_probes])
tre_genes <- unique(annotation$mgi_symbol[annotation$affy_probeset %in% tre_probes])

# subset to unambiguously TRE genes.
tre_genes <- tre_genes[!tre_genes %in% non_tre_genes]
gr_tre_genes <- gr_paper_tse_calls$symbol[gr_paper_tse_calls$call=="TRE"]
length(tre_genes)
## [1] 7116
length(gr_tre_genes) ## [1] 6611
length(intersect(gr_tre_genes, tre_genes))
## [1] 6281
```

With this very quick script we can easily re-indentify 95% of the genes reported as TSE in the GR paper.
The analysis for the GR paper was performed with a set of python scripts, sqlite databases and involved several additional filtering steps.